The Supervisionary proof-checking kernel

### 2021-10-26 paper/prisc-2022.tex
prisc-2022: more work on bibliography, a few more language tweaks

Interactive theorem proving software is typically designed around a trusted proof-checking \deffont{kernel}, the sole system component capable of authenticating theorems.
Untrusted automation procedures reside outside of the kernel, and drives it to deduce new theorems via an API.
Kernel and untrusted automation are typically implemented in the same programming language---the ``meta-language''---usually some functional programming language in the ML family.
This strategy---introduced by Milner in his LCF proof assistant~\cite{Milner1972LogicFC}---is a reliability mechanism, aiming to ensure that any purported theorem produced by the system is indeed a legitimate entailment from the theory within the logic.

@@

Note that one interesting consequence of this style of implementation is the ability to provide concise specifications of Supervisionary's system interface functions.
Essentially, the Supervisionary kernel is a grand exercise in pointer manipulation, and as such our ABI specifications can be expressed as Hoare Triples, using Separation Logic as our assertion language.
Writing $\mathtt{h} \mapsto_{\mathtt{trm}} \mathtt{Application(l, r)}$ to assert that the handle $\mathtt{h}$ points-to a term application (of the term pointed-to by $l$ to the term pointed-to by $r$), and writing $\mathtt{out} \mapsto \mathtt{b}$ to assert that $\mathtt{out}$ points-to the Boolean value $\mathtt{b}$, we have:
\begin{gather*}
\{ h \mapsto_{\mathtt{trm}} \mathtt{Application(l, r)} \} \\
\mathtt{Term.Handle.IsApplication(h, out)} \\
\{ \mathtt{out} \mapsto \mathtt{True} \}
\end{gather*}
(Here, the triple $\{ P \} C \{ Q \}$ asserts that if the command $C$ executes in a state concordant with $P$ then the command succeeds and produces a state concordant with $Q$.)

@@

We now turn to speculation around potential uses of Supervisionary.
Exploring what follows is still a work-in-progress.

% note: explicitly mention polyglot theorem proving software?

\paragraph{Runtime verification}

@@

But: \emph{what happens if we blur the lines between Supervisionary's interfaces for system access and proof-checking?}

@@

Note that this idea shares some similarities with \deffont{proof carrying code}, wherein binaries are accompanied with (skeleton) proofs of their adherence to some policy, and these proofs checked by the operating system prior to execution.
However, the ideas sketched above generalise this: proofs can be generated dynamically, as the program executes, and could more aptly be called \deffont{proof generating code}.

@@

For example, the operational semantics and instruction decoding functionality of Wasm is clearly embeddable in HOL.
Using this, properties of a program $P$ to be executed under Supervisionary could be established statically, and registered with the kernel, perhaps interactively by a user or by another program executing before $P$ executes.
This theorem can then be used by $P$ in closing challenges from Supervisionary, issued at runtime, tying the correctness of $P$ to its execution.

>>>>

Interactive theorem proving software is typically designed around a trusted proof-checking \deffont{kernel}, the sole system component capable of authenticating theorems.
Untrusted automation procedures reside outside of the kernel, and drives it to deduce new theorems via an API.
Kernel and untrusted automation are typically implemented in the same programming language---the ``meta-language''---usually some functional programming language in the ML family.
This strategy---introduced by Milner in his LCF proof assistant~\cite{Milner1972LogicFC}---is a reliability mechanism, aiming to ensure that any purported theorem produced by the system is indeed entailed by the theory within the logic.

@@

Note that one interesting consequence of this style of implementation is the ability to provide concise specifications of Supervisionary's system interface functions.
Essentially, the Supervisionary kernel is a grand exercise in pointer manipulation, and as such our system call specifications can be expressed as Hoare Triples, using Separation Logic~\cite{DBLP:conf/lics/Reynolds02} as our assertion language.
Writing $\mathtt{h} \mapsto_{\mathtt{trm}} \mathtt{Application(l, r)}$ to assert that the handle $\mathtt{h}$ points-to a term application (of the term pointed-to by $l$ to the term pointed-to by $r$), and writing $\mathtt{out} \mapsto \mathtt{b}$ to assert that $\mathtt{out}$ points-to the Boolean value $\mathtt{b}$, we have:
\begin{gather*}
\{ h \mapsto_{\mathtt{trm}} \mathtt{Application(l, r)} \} \\
\mathtt{Term.Handle.IsApplication(h, out)} \\
\{ \mathtt{out} \mapsto \mathtt{True} \}
\end{gather*}
(Here, the triple $\{ P \} C \{ Q \}$ asserts that if the command $C$ executes in a state concordant with $P$ then the command succeeds and produces a state concordant with $Q$.)

@@

\vspace{\baselineskip}
We now turn to speculation around potential uses of Supervisionary.
Exploring what follows is still a work-in-progress.
\paragraph{Runtime verification}

@@

\vspace{\baselineskip}
But: \emph{what happens if we blur the lines between Supervisionary's interfaces for system access and proof-checking?}
\vspace{\baselineskip}

@@

Note that this idea shares some similarities with \deffont{proof carrying code}, wherein binaries are accompanied with (skeleton) proofs of their adherence to some policy, and these proofs checked by the operating system prior to execution~\cite{DBLP:conf/popl/Necula97}.
However, the ideas sketched above generalise this: proofs can be generated dynamically, as the program executes, and could more aptly be called \deffont{proof generating code}.
Processes and the Supervisionary kernel work together to come to a ``mutual understanding'' that the behaviour of the process is indeed concordant with the prevailing policy.

@@

For example, the operational semantics and instruction decoding functionality of Wasm is clearly embeddable in HOL~\cite{DBLP:conf/cpp/Watt18}.
Using this, properties of a program $P$ to be executed under Supervisionary could be established statically, and registered with the kernel, perhaps interactively by a user or by another program executing before $P$ executes.
This theorem can then be used by $P$ in closing challenges from Supervisionary, issued at runtime, tying the correctness of $P$ to its execution.
